---
title: "Ice Cover"
format: html
editor: visual
---

```{r}
library(here)
library(tidyverse)
library(beeswarm)
library(lubridate)
```

```{r}
meteorlogical <- read_csv(here("data", "ntl20_v6.csv"))

ice_cover <- read_csv(here("data", "ntl33_v7.csv"), na ="-999")
```

Create a visualization that compares ice duration across different lakes in the Madison Lake Area. Include both the actual ice duration values *and* some representation of summary statistics for each lake. Careful if you use a boxplot over a jitter or beeswarm - what values might be doubly represented? Can you find an argument that would help you remove those from the boxplot?

```{r}
ice_cover |> 
  ggplot(aes(x = lakeid, y = ice_duration)) + geom_beeswarm() +geom_boxplot(outlier.shape = NA, alpha = 0.1)

```

Find the mean ice cover duration in Madison Lake Area by year
(i.e., you\'ll only be grouping by the year variable here). Understanding
check: what are you finding the mean of here? Why are there multiple
observations for each year?)

Finding the mean ice cover duration for all the lakes in the area, grouped by year. Multiple observations for each year are at different sites.

```{r}
summary <- ice_cover |> group_by(year4) |> 
  summarize(mean_ice_cover = mean(ice_duration, na.rm = TRUE))
```

Create an exploratory visualization of mean ice cover duration by
year for the Madison Lake Area. Add a sentence or two below this
exploratory graph describing the overall trend(s) you observe.

```{r}
summary |> ggplot(aes(x = year4, y = mean_ice_cover)) + geom_point() + 
  labs(caption = "The mean ice duration has declined between 1850 and 2000", x= "year", y = "Mean Ice duration") +
  theme_minimal()
  theme(plot.caption = element_text(hjust=0.2, size = 2))

```

Find the mean air temperature (using the daily average air
temperature - adjusted) in Madison Lake Area **by year**
using **only observations from winter months (December, January,
February)**

```{r}

met_summary <- meteorlogical |>
  filter(month %in% c(12,1,2)) |> 
group_by(year4) |> 
  summarize(mean_air_temp = mean(ave_air_temp_adjusted))
```

Create an exploratory visualization of mean winter temperatures
for the Madison Lake Area. Add a sentence or two below this exploratory
graph describing the overall trend(s) you observe.

```{r}
met_summary |> 
  ggplot(aes(x = year4, y = mean_air_temp)) + geom_line() + 
  labs(caption = "Mean air temp has increased over time")+
  theme(plot.caption = element_text(hjust = 0.2))
                                                            

```

Join the mean winter air temperatures to the mean ice cover
duration data you found above

```{r}
joined <-full_join(summary, met_summary, by = "year4")
```

Create an exploratory scatterplot of mean winter temperatures
versus mean ice duration. Add a sentence or two below this exploratory
graph describing the overall trend(s) you observe.

```{r}
joined |> 
  ggplot(aes(y = mean_air_temp, x = mean_ice_cover )) + geom_point()
```

Explore thaw dates since 1970. Create a subset of the ice cover
duration data since 1970 (ok to include 1970), then convert the
`ice_off` column to a `Date`. Use
`lubridate::yday()` to pull the numeric *day of the
year* from that Date. Create an exploratory plot of the ice off day
(numeric day-of-the-year) from 1970 - present. Add a sentence or two
below this exploratory graph describing the overall trend(s) you
observe.

```{r}
subset <- ice_cover |> 
  filter(year4 >= 1970) |> 
  mutate(ice_off = mdy(ice_off)) |> 
  mutate(ice_off_day = yday(ice_off))
```

```{r}
ggplot(data = subset, aes(x = year4, y = ice_off_day)) + geom_point() + labs(x = "year", y = "Day of year ice off", caption = "It appears that over time, the ice is more likely to melt earlier in the year")
```
